Gsmodutils: A python based framework for test driven genome scale metabolic model development

### 2017-11-29 article/article.tex
Added references and updated figure

@@

\begin{abstract}
\textbf{Motivation:} Genome scale metabolic models are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady state phenotypic behaviour.
Models of this scale can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models, little methodology is suggested for the management of continually curated large scale models.
This is quickly becoming an issue for large scale industrial biotechnology projects, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:} Taking advantage of python and cobrapy integration, we have developed the \textit{gsmodutils} modelling framework.
As a fully documented python API and series of command line utilities, gsmodutils places an emphasis on test driven design of models through easy to write test cases in json or python.
Crucially, different substrate conditions are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, hopefully, minimising error between model versions.
Furthermore, a core feature of the software is to allow users to easily create shareable, self contained, docker images that can be run in cross platform environments.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from http://github.com/gsmodutils \\
\textbf{Contact:}{james.gilbert@nottingham.ac.uk}
\end{abstract}

\section{Introduction}
Constraints based modelling for biological systems has become a mainstay of systems biology for several decades.
More recently, the structural approach has become a popular method for modelling steady state behaviour of large, metabolic networks.
Such large scale reconstructions are often refereed to as genome scale models (GSMs), as the processes is significantly aided throught the advent of genome sequencing.

@@

As a consequence, a significant amount of work has gone into the management of genome scale models.
The BiGG models database \cite{king2015bigg}, for example, exists to provide a standardised repository of validated models that can be shared and reused.
However, little focus is placed upon the collaborative design aspect of such models with few mechanisms existing to store the \textit{model delta} based tools provided in our software.
Similarly, the MetaNetX \cite{moretti2016metanetx} system exists to provide a standardised namespace and toolchain for GSMM analysis.
However, such tools often make it difficult to understand the design decisions made by the initial model authors.

Furthermore, the proliferation of tools available to users is significant.
This covers huge variety of software platforms including the COBRA toolbox in Matlab \cite{schellenberger2011quantitative},
ScrumPy and CobraPy in Python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux}, SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, there are often few options to make analysis easy to replicate.

@@

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential hetrologous pathways as well as providing predictions for flux distributions.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
However, significant changes to wild-type strains must be tracked and compared to allow models to remain relevant.

Similarly, many conventional applications of genome scale models in systems biology have often suffered from unnecessary replication of work due to a lack of standards.
For example, there are now three independently developed models of \textit{Clostridium acetobutylicum} \cite{mcanulty2012genome, dash2014capturing, yoo2015quantitative}, an organism used in the production of solvents for over a century.
These models all exist to solve similar biological problems, however there is disagreement over fundamental biochemical properties of these anaerobic organisms particularly with the focus on redox balancing \cite{dash2014capturing}.
Indeed, such models also include updates based on improved genome annotations and the inclusion of fluxomic, transcriptomic and metabolomic characterisations \cite{yoo2015quantitative}.

% Perhaps more worryingly, from an industrial perspective, is that the use of models in bioproduction has not escaped regulatory notice \cite{}.
Furthermore, as models are often used to explain issues with online production and facilitate the optimisation of bioprocessing, it stands to reason that model correctness should attract oversight.
Consequently, methods are needed to ensure that the models themselves are tightly version controlled.
In this paper, we argue for the use of a form of \textit{test driven} model of development, a conceptual approach borrowed from software engineering.

@@

The design phase of industrial synthetic biotechnology, due to historical and regulatory convention, often follows a conventional waterfall model for development of genome scale models.
This is borne out of a number significant constraints in the development of genome scale models.
The waterfall approach for model is characterised in Figure \ref{fig:methodology_approaches} where requirements for a model are captured at an early stage and are hard to refine during the production process.

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady state phenomena still provides a good understanding of systemic properties \cite{o2015using}.

Constraints based models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.

Such approaches often leave models with missing reactions \cite{}, incorrect gene-reaction rules \cite{} or with pathways based on gap filling methods that add reactions that may not be present within the organism \cite{benedict2014likelihood}.
Genome annotation is frequently updated and models are often corrected ``on the fly'' when attempting to understand specific natural phenomena.

@@

Consequently, models undergo significant manual annotation which can lead to errors.
In this work we advocate a test driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve core research goals that are dynamic in response to the changes of a project.
In order to achieve this goal, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that while many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions, their is a significant gap in terms of the reliability and re-usability of models.
The following sections provide an overview of the gsmodutils software framework.
Gsmodutils aims to provide a basis for test driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence.png}
  \label{fig:strain_hered}
  \caption{Examples of gsmodutils hierarchy for reuse and future compatibility of designs}
\end{figure}

@@

\section{Discussion}
When research projects end it all to is common for important large models to be published and become relics lost within the literature forgotten to all but the most dedicated of individuals.
Furthermore, as GSMs grow in terms of the metabolism the contain as well as the biological problems they are used to solve, problems with annotation and curation naturally accumulate as a product of human error.
We have presented a framework with a number of features taken from the software development world specifically designed to improve collaboration and minimise such error.
However, it is important to stress the difference between defined behaviour expected from pre-written test cases and novel predictions made by a model.
Indeed, a core objective of this framework is to ensure that good practices are followed in model development that help scientists to better trust the results discovered by their models.
In an ideal world, we would envision a methodology such as our becoming a pre-requisite for GSMs to pass peer review.

>>>>

%PREPRINT/MANUSCRIPT OPTIONS
\usepackage{setspace}
\usepackage{lineno}
\linenumbers
\doublespacing

@@

\begin{abstract}
\textbf{Motivation:} Genome scale metabolic models are increasingly important for systems biology and metabolic engineering research as they are capable of simulating complex steady state phenotypic behaviour.
Constraints based models of this form can include thousands of reactions and metabolites, with many crucial pathways that only become activated in specific simulation settings.
However, despite their widespread use, power and the availability of tools to aid with the construction and analysis of large scale models little methodology is suggested for the management of continually curated large scale models.
This is quickly becoming an issue for industrial systems and synthetic biotechnology applications, which require good quality reusable models integral to the design, build and test cycle.
\\
\textbf{Results:} Taking advantage of python and cobrapy integration, we have developed the \textit{gsmodutils} modelling framework.
As a fully documented python API and series of command line utilities, gsmodutils places an emphasis on test driven design of models through easy to write test cases in json or python.
Crucially, different substrate conditions are configurable allowing users to examine how different designs or curation impact a wide range of system behaviours, hopefully, minimising error between model versions.
Furthermore, a core feature of the software is to allow users to easily create shareable, self contained, docker images that can be run in cross platform environments.\\
\textbf{Availability:} The software framework described within this paper is open source and freely available from http://github.com/gsmodutils \\
\textbf{Contact:}{james.gilbert@nottingham.ac.uk}
\end{abstract}

\section{Introduction}
Stoichiometric constraints based modelling for biological systems has been a mainstay of systems biology for several decades \cite{fell1986fat, varma1994stoichiometric}.
Given its flexibility, low barrier to entry and requirement only on minimal knowledge regarding the stoichiometry of metabolic networks this structural approach has become an extremely popular method for modelling steady state behaviour of large, biochemical networks \cite{kauffman2003advances}.
Such large scale reconstructions are often refereed to as genome scale models (GSMs), as the processes is significantly aided through the advent of relatively inexpensive genome sequencing \cite{o2015using, land2015insights}.

@@

As a consequence, a significant amount of work has gone into the management of genome scale models.
The BiGG models database \cite{king2015bigg}, for example, exists to provide a standardised repository of validated models that can be shared and reused.
However, little focus is placed upon the collaborative design aspect of such models with few mechanisms existing to store the \textit{model delta} based tools provided in our software.
Similarly, the MetaNetX \cite{moretti2016metanetx} system exists to provide a standardised namespace and toolchain for GSM analysis.
However, such tools often make it difficult to understand the design decisions made by the initial model authors.

Furthermore, as with many areas of bioinformatic study the proliferation of available computational tools has become somewhat vast.
This covers huge variety of software platforms including the COBRA toolbox in Matlab \cite{schellenberger2011quantitative}, ScrumPy and CobraPy in Python \cite{poolman2006scrumpy, ebrahim2013cobrapy} with additional tools and libraries such as cameo \cite{cardoso2017cameo}, OptFlux in Java \cite{rocha2010optflux}, SurreyFBA \cite{gevorgyan2010surreyfba}.
Whilst most of these tools are Open Source and follow standards, such as SBML \cite{finney2003systems}, there are often few options that make analysis easy to replicate.

@@

For example, many tools such as RetroPath \cite{carbonell2013retropath}, XTMS \cite{carbonell2014xtms} and GEM-Path \cite{campodonico2014generation} suggest thousands of potential hetrologous pathways as well as providing predictions for flux distributions.
Many of these tools significantly increase the value of genome scale models, for example by coupling commodity production to an organism's growth \cite{feist2010model}.
These tools all suggest major changes to wild-type strains must be tracked and compared to allow models to remain relevant.
In effect, mechanisms are required to related modified models to modified test and production strains.

Similarly, many conventional applications of genome scale models in systems biology have often suffered from unnecessary replication of work due to a lack of standards.
For example, there are now many independently developed models of \textit{Clostridium acetobutylicum} \cite{lee2008genome, senger2008genome, mcanulty2012genome, dash2014capturing, yoo2015quantitative}, an organism used in the production of solvents for over a century.
These models all exist to solve similar biological problems some being updates to the inital base models. 
However, there has been disagreement over fundamental biochemical properties of these anaerobic organisms, notably with the focus on redox balancing \cite{dash2014capturing}.
Indeed, such models also include updates based on improved genome annotations and the inclusion of fluxomic, transcriptomic and metabolomic characterisations \cite{yoo2015quantitative}.

Many of the results presented in original sources are difficult to compare due to a lack of standardisation of reference identifiers for reaction names.
Where models are shared they are often in non-standard spreadsheet formats, rather than SBML models.
Indeed, even in the case of valid SBML models being made available at the time of publication few details are given as to how to run such models.

% Perhaps more worryingly, from an industrial perspective, is that the use of models in bioproduction has not escaped regulatory notice \cite{}.
Furthermore, as models are often used to explain issues with online production and facilitate the optimisation of bioprocessing, it stands to reason that model correctness should attract oversight.
Consequently, methods are needed to ensure that the models themselves are tightly version controlled.
In this paper, we argue for the use of a form of \textit{test driven} model of development, a conceptual approach borrowed from software engineering.
Given the availability of widespread open source virtualisation tools that allow the standardisation of platforms \cite{merkel2014docker}, modellers have the opportunity to share highly reproducible research \cite{chamberlain2014using}.

@@

The design phase of industrial synthetic biotechnology, due to historical and regulatory convention, often follows a waterfall model for development of genome scale models.
This is borne out of a number significant constraints in the development of genome scale models.
The waterfall approach for model is characterised in Figure \ref{fig:methodology_approaches} where requirements for a model are captured at an early stage and are hard to refine during the production process.

In contrast, an \textit{agile} methodology for the development of models places the focus on adapting work to new requirements \cite{martin2002agile}.
Such an approach best fits genome scale models because they are rarely created to investigate individual processes and, instead, capture the complexity of large systems.
Constraints based models are intrinsically related to available genome annotations.
Such annotations rely heavily on automated matches to related species, with the characterisations of individual genes often being left to a few of critical interest \cite{seemann2014prokka}.
This modelling formation costs an in depth understanding of dynamic behaviour. 
However, capturing steady state phenomena still provides a good understanding of systemic properties \cite{o2015using}.

As such approaches often leave models with missing reactions, incorrect gene-reaction rules \cite{thiele2010protocol} or with pathways based on gap filling methods that add reactions that may not be present within the organism \cite{benedict2014likelihood}.
When attempting to understand specific natural phenomena, genome annotations are frequently updated and models are often corrected in an \textit{ad hoc} manner.

@@

Therefore, models undergo significant manual annotation - a process which naturally leads to errors.
In this work we advocate a test driven approach to model development highlighted in Figure \ref{fig:test_driven}.
Here, the model is changed to achieve core research goals that are dynamic in response to the changes of a project.
In order to meet this objective, validation criteria for a model such as growth conditions or the impact of gene knock outs, should be formally set.
When a model is changed, all such validation criteria should be retested to ensure that models do not regress to previous states.

@@

We feel that while many of the current software tools for genome scale models do an excellent job of facilitating answers to crucial research and design questions.
However, their is a major gap in terms of the reliability and re-usability of models due to a lack of standardisation and software tools to aid such processes.
The following sections provide an overview of the gsmodutils software framework.
Gsmodutils aims to provide a basis for test driven, version controlled agile model development.
All software is open source and is designed to be interoperable with platforms widely used in the domain of constraints based modelling.

@@

\begin{figure}
  \includegraphics[width=\textwidth]{inheritence.png}
  \label{fig:strain_hered}
  \caption{Examples of gsmodutils design inheritance.
  Each design stores the delta between the wild-type base model, any parents and the changes to constraints the design contains.
  In the example presented above a hetrologous production pathway is combined with a reusable set of knock-outs.
  In practice, ideally, these designs should relate to real constructs and strains evaluated by wet lab biologists.
  }
\end{figure}

@@

\section{Discussion}
In order to facilitate the sharing and dissemination of research good standards and software are required \cite{}.
Naturally a great deal of effort has gone in to producing high quality systems and synthetic biology standards \cite{}.
Furthermore, when research projects end it all to is common for important large models to be published and become relics lost within the literature forgotten to all but the most dedicated of individuals.
As GSMs grow in terms of the metabolism the contain as well as the biological problems they are used to solve, problems with annotation and curation naturally accumulate as a product of human error.
Software that facilitates actively improving how researchers develop and apply models to new phenomena is required.

We have presented a framework with a number of features taken from the software development world specifically designed to improve collaboration and minimise such error.
However, it is important to stress the difference between defined behaviour expected from pre-written test cases and novel predictions made by a model.
Indeed, a core objective of this framework is to ensure that good practices are followed in model development that help scientists to better trust the results discovered by their models.
In an ideal world, we would envision a methodology such as our becoming a pre-requisite for GSMs to pass peer review.
